Enhancing the EEG signals visualization

### 2023-07-17 article/article.tex
For (if any) god's sake! Almost there! Just lacks the explanation of the sourcecode!

@@

	\section{Introduction}
		\par Electroencephalogram (EEG) data is widely used in neuroscience research to study brain activity. EEG records electrical signals generated by brain neurons and reflects cognitive processes, making it a valuable tool in various applications, including Brain-Computer Interfaces (BCI) and clinical diagnosis of neurological disorders \cite{8937083}. Analyzing EEG data, however, poses challenges due to its complexity, such as a vast number of emitting sources (neurons) and potential electromagnetic interferences from external sources like cell phones, electric motors, and magnetic fields from wires \cite{8937083}. To facilitate EEG data interpretation and visualization, it is crucial to develop effective and efficient visualization techniques that allow researchers to gain valuable insights from the data. This paper proposes a novel visualization method for EEG data, combining line graphs and energy visualizations with a display of the electrode positions in the 10-20 system \cite{sistema10-20}. The proposed visualization will be interactive, allowing users to select subjects and stimuli and explore specific signal details. The goal is to provide researchers with an informative and comprehensive view of EEG data to support their investigations.

	\section{Data Acquisition}
		\par To demonstrate the proposed visualization technique, EEG data will be acquired from an existing dataset. The chosen dataset is the "Open access database of EEG signals recorded during imagined speech" \cite{10.1117/12.2255697}. This dataset contains EEG recordings during the imagined pronunciation of vowels and commands, as well as EEG and audio recordings during the actual pronunciation. The dataset includes recordings from 15 Argentinian volunteers between the ages of 24 and 28, with an equal representation of males and females \cite{10.1117/12.2255697}. Although this dataset serves as the initial data source for the proposed visualization, our future research intends to develop a custom EEG database in collaboration with local health institutions.

	\section{Database Structure}
		\par Each row in the dataset corresponds to a subject-stimuli-modality set, and the data is arranged as follows: EEG signals from different electrodes, modalities (imagined or pronounced), and stimulus codes representing different categories of stimuli (e.g., letters and commands). The modalities distinguish between imagined and pronounced speech, while the stimulus codes identify specific categories of stimuli. Figure \ref{fig:visu04} shows a sample of the initial rows in the database, and Figure \ref{fig:visu05} displays the last rows.

@@

	\section{Pre-processing}
		\par To facilitate data manipulation and visualization, pre-processing techniques are applied to the EEG data. The data is grouped and organized to ensure seamless interactions and efficient representation. Figure \ref{fig:visu06} demonstrates the pre-processed and grouped data, ready for visualization.

@@

	\section{Case Studies}
		\par The proposed visualization method will be evaluated through case studies using real EEG data. The case studies will focus on different subjects and stimuli to demonstrate the effectiveness and utility of the proposed visualization in various scenarios.

		\subsection{Visualization Inspired by EPviz \cite{currey2023epviz}}
			\par The proposed visualization method draws inspiration from EPviz, which effectively uses line graphs to visualize EEG data \cite{currey2023epviz}. EPviz visualizes EEG signals captured by all electrodes, providing a general overview of the data. However, the visualization lacks detailed views, which may limit the ability to discern specific patterns and features.

@@

		\subsection{Visualization Inspired by \cite{9098189}}
			\par Another relevant methodology uses topological color maps to indicate signal energy at each sensor, offering insights into brain activity \cite{9098189}. The visualization in Figure \ref{fig:visu01} provides an informative representation of regions with intense brain activity during the corresponding time period. However, the interpolation of measured values may lead to potential misinterpretations of signal intensities according to gradients.

@@

		\subsection{Visualization Inspired by \cite{8937083}}
			\par This methodology shown in Figure \ref{fig:visu02} simplifies the visualization to display brain activity at four different moments \cite{8937083}. However, similar to the previous visualization, the gradient issue persists, potentially leading to incorrect interpretations.

@@

	\section{Proposal}
		\par The proposed visualization method aims to overcome the limitations of existing techniques. It includes line graphs and energy visualizations combined with a non-gradient representation of electrode positions in the 10-20 system. The interactive features allow users to select subjects and stimuli, explore specific signal details, view the signal's origin, and amplify the displayed curves. Additionally, the visualization will include a bar showing the total energy of the displayed signal interval. The use of high-contrast colors will aid in perceptual differentiation. Figures \ref{fig:g3714} and \ref{fig:g3762} illustrate the general and specific views of the proposed visualization.

@@

	\section{Implementation}
		\par The visualization was implemented using the Python3 \cite{Python3} programming language due to its extensive range of data visualization and manipulation libraries. It is important to note that the final result differs slightly from the proposal, however, this does not compromise the quality of the final product, as can be seen in Figure \ref{fig:screenshot01} and \ref{fig:screenshot02}.

@@

@@

	\section{References}
	\bibliographystyle{ACM-Reference-Format}
	\bibliography{../presentation/bibliography.bib}

>>>>

\input{preamble.tex}

@@

	\section{Introduction}
		\par Electroencephalogram (EEG) data is widely used in neuroscience research to study brain activity. EEG records electrical signals generated by brain neurons during a task making it a valuable tool in various applications, including Brain-Computer Interfaces (BCI) and clinical diagnosis of neurological states \cite{8937083}. Analyzing EEG data, however, poses challenges due to its complexity, such as a vast number of emitting sources (neurons) and potential electromagnetic interferences from external sources like cell phones, electric motors, and magnetic fields from wires \cite{8937083}. To facilitate EEG data interpretation and visualization, it is crucial to develop effective and efficient techniques that allow researchers to gain valuable insights from the data. This paper proposes a visualization method for EEG data, combining line graphs and energy visualizations with a display of the electrode positions in the 10-20 system \cite{sistema10-20}. The proposed visualization will be interactive, allowing users to select subjects and stimuli and explore signal details. The goal is to provide researchers with an informative and comprehensive view of EEG data to support their investigations.

	\section{Data Acquisition}

		\par To demonstrate the proposed visualization technique, EEG data was acquired from an existing dataset. The chosen dataset is the "Open access database of EEG signals recorded during imagined speech" \cite{10.1117/12.2255697}. This dataset contains EEG recordings during the imagined pronunciation of vowels and commands, as well as EEG and audio recordings during the actual pronunciation. The dataset includes recordings from 15 Argentinian volunteers between the ages of 24 and 28, with an equal representation of males and females \cite{10.1117/12.2255697}.\newline

		\par As shown in Figure \ref{fig:sistema10-20}, the data was acquired using the 10-20 system \cite{sistema10-20} which consist of electrodes spaced by 20\% of the head size and a margin of 10\% in relation to \textit{nasion}, \textit{inion} and to the preauricular point. In this specific database only the sensors $F3$, $F4$, $C3$, $C4$, $P3$ and $P4$ was used.

		\begin{figure}[h]
			\centering
			\includegraphics[width=\linewidth]{../presentation/images/sistema10-20}
			\caption{system 10-20: The electrodes are evenly separated across the scalp of the subject. The A1 sensor is the preauricular point. Source: \cite{sistema10-20}}
			\label{fig:sistema10-20}
		\end{figure}

		\par Although this dataset serves as the initial data source for the proposed visualization, our future research intends to develop a custom EEG database in collaboration with local health institutions.

	\section{Database Structure}
		\par Each row in the dataset corresponds to a subject-stimuli-modality set, and the data is arranged as follows: EEG signals from different electrodes, modalities (imagined or pronounced), and stimulus codes representing different categories of stimuli (e.g., letters and commands). The modalities distinguish between imagined and pronounced speech, while the stimulus codes identify specific categories of stimuli. Figure \ref{fig:visu04} shows a sample of the initial rows in the database, and Figure \ref{fig:visu05} displays the last rows.\newline

		\par Each line consist in a series of numbers organized as follows:
		\begin{itemize}
			\item F3 - 0:4095
			\item F4 - 4096:8191
			\item C3 - 8192:12287
			\item C4 - 12288:16383
			\item P3 - 16384:20479
			\item P4 - 20480:24575
			\item modality - 24576:24576
			\item estimuli - 24577:24577
			\item artifact - 24578:24578
		\end{itemize}

		\par Modality can be 1 for imagined speech or 2 for pronounced speech. Stimuli assumes 1: A, 2: E, 3: I, 4:O and 5:U and 6:\textit{Arriba}, 7: \textit{Abajo}, 8: \textit{Adelante}, 9: \textit{Atrás}, 10: \textit{Derecha}, 11: \textit{Izquierda}.

@@

	\section{Preprocessing}
		\par To facilitate data manipulation and visualization, pre-processing techniques are applied to the EEG data. The data is grouped and organized to ensure better interactions and efficient representation. Figure \ref{fig:visu06} demonstrates the pre-processed and grouped data, ready for visualization.

@@

	\section{Case Studies}
		\par The proposed visualization method drew inspiration from case studies that utilized real EEG data. These case studies focused on various subjects and stimuli, showcasing their effectiveness and utility. Accordingly, certain features will be incorporated, while others will be discarded.

		\subsection{Visualization Inspired by \cite{currey2023epviz}}
			\par EPviz uses line graphs to visualize EEG data \cite{currey2023epviz} and provide a general overview of the data. The visualization by lines is efficient because it shows the high and low frequency components. In the example shown in Figure \ref{fig:epviz00}, the EPViz program visualizes the signals captured by all the electrodes, providing a good overview. However, it lacks the detailed view that would allow a more accurate analysis of the displayed curves.

@@

		\subsection{Visualization Inspired by \cite{9098189}}
			\par Another relevant methodology uses topological color maps to indicate signal energy at each sensor. This visualization approach shown in Figure \ref{fig:visu01} provides an informative representation of regions with intense brain activity during the corresponding time period. However, the view makes a interpolation of measured values leading to potential misinterpretations of signal intensities according to gradients.

@@

		\subsection{Visualization Inspired by \cite{8937083}}
			\par The methodology shown in Figure \ref{fig:visu02} simplifies the visualization to display brain activity at four different moments. However, similar to the previous visualization, the gradient issue persists, potentially leading to incorrect interpretations.

@@

	\section{Proposal}
		\par The proposed visualization method includes line graphs and energy visualizations combined with a non-gradient representation of electrode positions in the 10-20 system. The interactive features allow users to select subjects and stimuli, explore specific signal details, view the signal's origin, and scale up or down the displayed curves. Additionally, the visualization will include bars showing the total energy of the displayed signal interval. The use of high-contrast colors will aid in perceptual differentiation. Figures \ref{fig:g3714} and \ref{fig:g3762} illustrate the general and specific views proposed.

@@

		\par Both the \textbf{subject and stimulus selection} and the checkbox are integral components of the \textbf{visualization cycle} \cite{munzner2014visualization}. When modifying the type of vision, stimuli and/or the analyzed subject, the view must promptly respond by updating the displayed data.\newline

		\par The use of \textbf{high contrast} colors aids in easily locating the origin of the data \cite{spence2014information}, preventing the common problem of differentiating data when very similar colors are employed.\newline

		\par The \textbf{magnification} feature enables a detailed view of each signal \cite{tufte1983visual}, facilitating the visualization of information linked to low and high-frequency signals. This provides the identification of \textbf{objects according to their "silhouette"} \cite{tufte2006beautiful}.\newline

		\par It's worth noting that the attributes utilized in the proposal have a \textbf{sequential} \cite{kirk2012data} order for the amplitude and frequency of the signals over time (line graphs), while simultaneously being \textbf{categorical} \cite{kirk2012data} because the frequency analysis can reveal valuable insights about the subject being analyzed.\newline

		\par The energy attribute (vertical bar on the side of the line graph) is \textbf{quantitative}.\newline

		\par Regarding \textbf{categorical} attributes, the region to the left of the graphs also deserves attention. This section indicates which brain regions are emitting specific signals, allowing, for example, the inference of the most active brain regions during thought execution and/or action.\newline

		\par In terms of \textbf{semiotic categories} \cite{santaella2017semiotica} it is possible to highlight:
		\begin{itemize}
			\item firstness: Colors that stand out and immediately call the user's attention.
			\item secondness: The action/reaction applied to controls and images.
			\item thirdness: The consequent interpretation of the image presented as well as the reasoning necessary for understanding what is shown.
		\end{itemize}

	\section{Implementation}
		\par The visualization was implemented using the Python3 \cite{Python3} programming language due to its extensive set of data visualization and manipulation libraries. It is important to note that the final result differs slightly from the proposal, however, this does not compromise the quality of the final product, as can be seen in Figure \ref{fig:screenshot01} and \ref{fig:screenshot02}.

@@

			\lstinputlisting[language=Python]{code/test.py}

@@

	\bibliographystyle{ACM-Reference-Format}
	\bibliography{../presentation/bibliography.bib}
